Verifying Spatial Properties of Array Computations

### 2017-04-18 oopsla17/study.tex
remove spurious comma
some editing in study

\subsection{Methodology}
%
\noindent
We constructed a corpus of eleven scientific computing packages
written in Fortran ranging in size and scope:
%
\begin{enumerate}[leftmargin=1.5em]
\item The Unified Model (UM) developed by the UK Met Office
for weather modelling~\cite{um};
\item E3ME, a mixed economic/energy impact predication
  model~\cite{RePEc:aen:journl:2006se-a12}

@@

\begin{center}
\begin{tabular}{l||l|p{0.28\linewidth}|l}
  Property   & Shorthand & Classifications (of RHS pattern) & Example \\ \hline
%%%%%%
\multirow{2}{*}{Contiguity} & (\textsf{contig}) & Contiguous  &
                                                                \fortran{a(i) + a(i+1) + a(i+2)} \\
  & (\textsf{disjoint}) & Non-contiguous & \fortran{a(i) + a(i+2)} \\ \hline
%%%%%%
%\multirow{2}{*}{Shape} & (\textsf{rect}) & (Hyper)rectangle &
% \fortran{a(i,j) + a(i+1,j) + a(i,j+1) + a(i+1,j+1)} \\
%           & (\textsf{sumRect}) & Composed (hyper)rectangles &
%  \fortran{a(i,j) + a(i,j-1) + a(i-1,j) + a(i+1,j) + a(i,j+1)} \\ \hline
%%%%%%
  \multirow{2}{*}{Reuse} & (\textsf{readonce}) & Unique subscripts
                                         & \fortran{b(i) = a(i) + a(i+1)} \\
  & (\textsf{mult}) & Repeated subscripts &
 \fortran{b(i) = a(i)  + a(i)} \\ \hline
  \multirow{2}{*}{Positioning} & (\textsf{origin}) & Includes origin
& \fortran{a(i)} or \fortran{a(i+1, j)} (in $2^{\textit{nd}}$
  dimension) \\
             & (\textsf{straddle}) & Within distance 1 of origin &
\fortran{a(i+1,j) + a(i-1,j)} \\
             & (\textsf{away}) & Away from the origin
                                          & \fortran{a(i+2), a(i+3)}
\end{tabular}
\end{center}
%
%To assess the hypothesis that
%arrays are mostly read in a pattern that includes the immediate neighbours
%to the ``origin'', we categorised the position of subscript patterns for each
%right-hand side, in each dimension: \\
%
%\noindent
%
Finally, we categorise the relationship between the left-hand side
and right-hand side, in terms of the use of induction variables.
The two sides are \emph{consistent} if the same induction variables
appear in each side, used for the same dimension. This is
weakened to a \emph{permutation} if the roles of the induction
variables changes. This is weakened further if the left-hand side
induction variables are either
a subset or superset of the induction variables on the right-hand
side. Otherwise, the two sides are seen as inconsistent:
%
\begin{center}
\begin{tabular}{l|l}
  %Shorthand &
  Classification & Example \\ \hline
  %(\textsf{const}) &
  Consistent & \fortran{a(i, j) = b(i, j) +
                                  b(i+1,j+1)} \\
  %(\textsf{perm})
  Permutation & \fortran{a(i, j) = c(j, i)} or \fortran{a(i,
                                                               0) =
                                 b(0, i)} \\
  %(\textsf{LHSsub}) &
  LHS subset & \fortran{a(i) = b(i, j) + b(i, j-1)}
  \\
  %(\textsf{LHSsup}) &
  LHS superset & \fortran{a(i, j) = b(i)} \\
  %(\textsf{inconst}) &
  Inconsistent & \fortran{a(i) = b(j)}
\end{tabular}
\end{center}
%
If an assignment is not inconsistent, but is one of the other four
categories, we refer to it as *consistent.

@@

& \textbf{\# *consistent}

@@

indexed by induction variables, and a right-hand side comprising

@@

have a static pattern on the left- and right-hand side that is either
a neighbour or affine, and $61.79\%$ we classify as \emph{stencils},

@@

Most stencils are contiguous, but there is some
non-contiguity. Most stencils have matching induction
variables on the left and right-hand sides but again there
are non-trivial amounts of the other kinds of relationship.
%%
\begin{enumerate}[resume]
\item Many array computations read from each particular index just
  once.
\end{enumerate}
%
We classified right-hand sides comprising neighbourhood (or
affine) subscripts and subclassified this by whether array subscript terms were
all unique or whether there were some repeats.
%
\begin{center}
\begin{tabular}{l|S[table-format=4]S[table-format=2.2]S[table-format=4]}
\textbf{RHS form and reuse} & \textbf{Number} & \% \\ \hline
RHS neighbour/affine unique subscripts      & 78349     & 72.03\%  \\\hline
RHS neighbour/affine repeated subscripts   & 10228     & 9.40\%  \\\hline
other                             & 20196     & 18.57\%  %\\\hline\hline
%Total                             & 108773    &
\end{tabular}
\end{center}
%

Thus most array computations have access patterns with
unique use of array subscripts, confirming Hypothesis 5.

>>>>

\subsection{Methodology}
%
\noindent
We constructed a corpus of eleven scientific computing packages
written in Fortran ranging in size and scope:
%
\begin{enumerate}[leftmargin=1.76em]
\item The Unified Model (UM) developed by the UK Met Office
for weather modelling~\cite{um};
\item E3ME, a mixed economic/energy impact predication
  model~\cite{RePEc:aen:journl:2006se-a12}

@@

\begin{center}
\begin{tabular}{l||l|p{0.28\linewidth}|l}
  Property   & Shorthand & Classifications (of RHS pattern) & Example \\ \hline
%%%%%%
\multirow{2}{*}{Contiguity} & (\textsf{contig}) & Contiguous  &
                                                                \fortran{a(i) + a(i+1) + a(i+2)} \\
  & (\textsf{disjoint}) & Non-contiguous & \fortran{a(i) + a(i+2)} \\ \hline
%%%%%%
%\multirow{2}{*}{Shape} & (\textsf{rect}) & (Hyper)rectangle &
% \fortran{a(i,j) + a(i+1,j) + a(i,j+1) + a(i+1,j+1)} \\
%           & (\textsf{sumRect}) & Composed (hyper)rectangles &
%  \fortran{a(i,j) + a(i,j-1) + a(i-1,j) + a(i+1,j) + a(i,j+1)} \\ \hline
%%%%%%
  \multirow{2}{*}{Reuse} & (\textsf{readonce}) & Unique subscripts
                                         & \fortran{b(i) = a(i) + a(i+1)} \\
  & (\textsf{mult}) & Repeated subscripts &
 \fortran{b(i) = a(i)  + a(i)} \\ \hline
  \multirow{2}{*}{Positioning} & (\textsf{origin}) & Includes origin
& \fortran{a(i)} or \fortran{a(i+1, j)} (in $2^{\textit{nd}}$
  dimension) \\
             & (\textsf{straddle}) & Within distance 1 of origin &
\fortran{a(i+1,j) + a(i-1,j)} \\
             & (\textsf{away}) & Away from the origin
                                          & \fortran{a(i+2), a(i+3)}
\end{tabular}
\end{center}
%
%To assess the hypothesis that
%arrays are mostly read in a pattern that includes the immediate neighbours
%to the ``origin'', we categorised the position of subscript patterns for each
%right-hand side, in each dimension: \\
%
%\noindent
%
Finally, we categorise the relationship between the left-hand side
and right-hand side, in terms of the use of induction variables.
The two sides are \emph{consistent} if the same induction variables
appear in each side, used for the same dimension. This is
weakened to a \emph{permutation} if the roles of the induction
variables changes. This is weakened further if the left-hand side
induction variables are either
a subset or superset of the induction variables on the right-hand
side. Otherwise, the two sides are seen as inconsistent:
%
\begin{center}
\begin{tabular}{l|l|l}
  %Shorthand &
  Classification & Sub Classification & Example \\ \hline
  %(\textsf{const}) &
  Consistent & Direct & \fortran{a(i, j) = b(i, j) +
                                  b(i+1,j+1)} \\
  %(\textsf{perm})
  & Permutation & \fortran{a(i, j) = c(j, i)} or \fortran{a(i,
                                                               0) =
                                 b(0, i)} \\
  %(\textsf{LHSsub}) &
  & LHS subset & \fortran{a(i) = b(i, j) + b(i, j-1)}
  \\
  %(\textsf{LHSsup}) &
  & LHS superset & \fortran{a(i, j) = b(i)} \\ \hline
  %(\textsf{inconst}) &
  \multicolumn{2}{l|}{Inconsistent} & \fortran{a(i) = b(j)}
\end{tabular}
\end{center}
%

@@

& \textbf{\# Consistent}

@@

indexed by induction variables, and an RHS comprising

@@

have a static pattern on the LHS and RHS that is either
a neighbourhood or affine, and $61.79\%$ are \emph{stencils},

@@

Most stencils are contiguous, but there is some
non-contiguity. Most stencils have matching induction
variables on the left and right-hand sides but again there
are non-trivial amounts of the other kinds of relationship.
%%
\begin{enumerate}[resume]
\item Many array computations read from each particular index just
  once.
\end{enumerate}
%
We classified right-hand sides comprising neighbourhood (or
affine) subscripts and subclassified this by whether array subscript terms were
all unique or whether there were some repeats.
%
\begin{center}
\begin{tabular}{l|S[table-format=4]S[table-format=2.2]S[table-format=4]}
\textbf{RHS form and reuse} & \textbf{Number} & \% \\ \hline
RHS neighbour/affine unique subscripts      & 78349     & 72.03\%  \\\hline
RHS neighbour/affine repeated subscripts   & 10228     & 9.40\%  \\\hline
other                             & 20196     & 18.57\%  %\\\hline\hline
%Total                             & 108773    &
\end{tabular}
\end{center}
%
Thus most array computations have access patterns with
unique use of array subscripts, confirming Hypothesis 5.
